\textsc{Atomic modeling of argon}\\ \footnotesize{\textsc{Project 5 --- FYS3150}}

### 2015-11-29 projects/project_5/molecular_dynamics/article/article.tex
fixed some typos and wonky sentences

\begin{abstract}
\end{abstract}
\tableofcontents
\section{Introduction}
\label{sec:introduction}
    Molecular dynamics (referred to as MD), deals with the time evolution of a
    system of particles, typically atoms or molecules. Being first explored in
    the late 1950's, it has had a wide range of applications, in fields as
    materials science, biochemistry and biophysics. It is based on the simple
    idea of considering a finite number of molecules and letting them interact
    for a fixed amount of time, which gives us an idea of how the system
    evolves.

@@

@@

    Initially, we are presented with the skeleton of a MD-program, and we are
    tasked with filling in the blanks. In this section, we walk through each of
    the steps that went into the implementation of the program, from skeleton
    to working program. The process can be summarized as

@@

    We want our system to conform to periodic boundary conditions. The reasons
    for doing this is that to a certain extent the system of finite size more
    closely resembles that of an infinite size. It is important that the
    bounding box which determines the boundary of the system is of sufficient
    size, so we do not get artifacts that arise from the fact that an atom can
    interact with itself.

@@

    Mathematically, the distribution function reads
    \begin{equation}
        \notag
        P(v_i)\, dv_i = \left( \frac{m_i}{2\pi k_BT} \right)^{1/2} e
    \end{equation}

@@

    In order to improve upon the fairly poor energy conservation of the
    Euler-Cromer method, we chose to implement the Velocity Verlet integrator.
    The algorithm is dependent on us having computed the forces for the first
    step and consists of three steps per time-step:
    \begin{align*}
        \vec{v}(t + \Delta t/2) &= \vec{v}(t) + \frac{\vec{F}(t)}{m}\frac{\Delta t}{2},\\
        \vec{r}(t + \Delta t) &= \vec{r}(t) + \vec{v}(t + \Delta t/2)\Delta t, \\
        \vec{v}(t + \Delta t) &= \vec{v}(t + \Delta t/2) + \frac{\vec{F}(t+\Delta t)}{m} \frac{\Delta t}{2}.
    \end{align*}
    We implement this algorithm in the \texttt{integrate} method in the
    \texttt{VelocityVerlet}-class. The method takes a system of atoms and a
    time step $\Delta t$ as arguments. We also introduce the private auxiliary
    variable \texttt{m\_first\_step}, which helps us takes care of the initial
    calculation of the forces.

@@

    We are also interested in the instantaneous temperature of the system, and
    the equipartition theorem \cite{equipar} can help us with this. The theorem
    relates the average energy of a system to its temperature and gives us the
    relation $\langle E_k \rangle = (3/2)k_bT$ which can be solved for
    temperature:
    \begin{equation}
        \notag
        T = \frac{2}{3} \frac{E_k}{nk_b}.
    \end{equation}
    Notice the removal of the brackets --- we are only interested in the
    \emph{instantaneous} temperature at any point in time. We also recall that
    since our units are scaled, Boltzmann's constant is equal to one.

@@

We initially decided to use the Velocity-Verlet integration method instead of
Euler-Cromer because of its good energy conservation. It is therefore of
interest to examine exactly what kind of impact this change of integration
method has for our simulations.  We now simulate a set of similar systems with
varying temperature and look at how the fluctuations in total energy scales
with the time step $\Delta T$. In order to do this we can consider the
standard-deviation of the total energy.

@@

We are now interested in determining what initial temperature $T_i$ is needed
in order to have a final temperature of $T$.  One way of examining this is by
looking at the ratio $T/T_i$. As we can see in \ref{fig:temperature_ratio} the
ratio stabilizes around $T/T_i = 0.5$. Solving this for $T_i$ tells us that if
we are interested in a final temperature of $T$, then we should chose our
initial temperature to be
\begin{equation}
    \notag
    T_i = 2T.
\end{equation}

@@

\begin{figure}[h]
    \centering \includegraphics[width=0.8\linewidth]{temperature_ratio.pdf}
    \caption[Temperature ratio]{Plotting the ratio between initial and
        instantaneous temperature for varying $T_i$. Here the initial
        temperatures ranges from 20 to 500 kelvin, with 48 intermediate steps.
    As we can see the ratio stabilizes around 0.5. The graph of an arbitrary
temperature has been highlighted in black.}
    \label{fig:temperature_ratio}
\end{figure}

@@

In order to measure the melting temperature of the system, we can use the Einstein relation which states the following:
\begin{equation}
    \notag
    \langle r^2(t) \rangle = 6Dt.
\end{equation}
$D$ is here the diffusion constant, $t$ is time and the \emph{mean square displacement} is calculated as
\begin{equation}
    \notag
    r_i^2(t) = \left| \vec{r}_i(t) - \vec{r}_i(0) \right|^2.
\end{equation}
\listoffigures

\begin{thebibliography}{9}

\bibitem{min_image}
    \url{https://en.wikipedia.org/wiki/Periodic_boundary_conditions#Practical_implementation:_continuity_and_the_minimum_image_convention}
\bibitem{equipar}
    \url{https://en.wikipedia.org/wiki/Equipartition\_theorem}
\bibitem{virial}
    \url{https://en.wikipedia.org/wiki/Virial_theorem#Time_averaging}
\end{thebibliography}

\end{document}

>>>>

\begin{abstract}
    In this project we explore the world of molecular dynamics. We are first
    presented with the skeleton of a working MD-program, and we are asked to
    implement the functionality we need as we go along. We look at the time
    evolution of a system of argon, with an initial face-centered cubic lattice
    configuration. The atoms are under the influence of the Lennard-Jones
    potential. We use the Velocity-Verlet as our time integrator of choice, and
    we compare this to the Euler-Cromer method in terms of energy conservation
    properties. We also compute the melting temperature of argon, by means of
    the Einstein relation and diffusion constant.
\end{abstract}
\tableofcontents
\listoffigures

\section{Introduction}
\label{sec:introduction}
    Molecular dynamics (referred to as MD), deals with the time evolution of a
    system of particles, typically atoms or molecules. Being first explored in
    the late 1950's, it has had a wide range of applications, in fields as
    materials science, biochemistry and biophysics. It is based on the simple
    idea of considering a finite number of molecules and letting them interact
    for a fixed amount of time, which gives us an idea of how the system
    evolves.

@@

    The construction of the program will be covered in section
    \ref{sec:building_a_working_md_program}, where some of the motivation
    behind each addition to the program will be discussed. In section
    \ref{sec:simulations} we employ our newly implemented program and look at
    some specific systems.

@@

    Initially, we are presented with the skeleton of a MD-program, and we are
    tasked with filling in the blanks. In this section, we walk through each of
    the steps that went into the implementation of the program, from skeleton
    to working program. The process can be summarized as following:

@@

    We want our system to conform to periodic boundary conditions. The reasons
    for doing this is that a finite system with periodic boundary condition
    more closely resembles that of an infinite size. It is important that the
    bounding box which determines the boundary of the system is of sufficient
    size, so we do not get artifacts that arise from the fact that an atom can
    interact with itself.

@@

    Mathematically, the distribution function reads
    \begin{equation}
        \notag
        P(v_i)\, dv_i = \left( \frac{m_i}{2\pi k_BT} \right)^{1/2} e^{-\frac{mv_i^2}{2k_BT}}dv_i
    \end{equation}

@@

    In order to improve upon the fairly poor energy conservation of the
    Euler-Cromer method, we chose to implement the Velocity Verlet integrator.
    The algorithm is dependent on us having computed the forces for the first
    step and consists of three steps per time-step:
    \begin{align*}
        \vec{v}(t + \Delta t/2) &= \vec{v}(t) + \frac{\vec{F}(t)}{m}\frac{\Delta t}{2},\\
        \vec{r}(t + \Delta t) &= \vec{r}(t) + \vec{v}(t + \Delta t/2)\Delta t, \\
        \vec{v}(t + \Delta t) &= \vec{v}(t + \Delta t/2) + \frac{\vec{F}(t+\Delta t)}{m} \frac{\Delta t}{2}.
    \end{align*}
    We implement this algorithm in the \texttt{integrate} method in the
    \texttt{VelocityVerlet}-class. The method takes a system of atoms and a
    time step $\Delta t$ as arguments. We also introduce the private auxiliary
    variable \texttt{m\_first\_step}, which helps us take care of the initial
    calculation of the forces.

@@

    We are also interested in the instantaneous temperature of the system, and
    the equipartition theorem \cite{equipar} can help us with this. The theorem
    relates the average energy of a system to its temperature and gives us the
    relation $\langle E_k \rangle = (3/2)k_BT$ which can be solved for
    temperature:
    \begin{equation}
        \notag
        T = \frac{2}{3} \frac{E_k}{nk_B}.
    \end{equation}
    Notice the removal of the brackets --- we are only interested in the
    \emph{instantaneous} temperature at any point in time. We also recall that
    since our units are scaled, Boltzmann's constant is equal to one.

@@

We initially decided to use the Velocity-Verlet integration method instead of
Euler-Cromer because of its good energy conservation. It is therefore of
interest to examine exactly what kind of impact this change of integration
method has for our simulations.  We now simulate a set of similar systems and
look at how the fluctuations in total energy scales with the time step $\Delta
T$. In order to do this we can consider the standard-deviation of the total
energy.

@@

We are now interested in determining what initial temperature $T_i$ is needed
in order to have a final temperature of $T$.  One way of examining this is by
looking at the ratio $T/T_i$. As we can see in figure \ref{fig:temperature_ratio} the
ratio stabilizes around $T/T_i = 0.5$. Solving this for $T_i$ tells us that if
we are interested in a final temperature of $T$, then we should chose our
initial temperature to be
\begin{equation}
    \notag
    T_i = 2T.
\end{equation}

@@

\begin{figure}[h]
    \centering \includegraphics[width=\linewidth]{temperature_ratio.pdf}
    \caption[Temperature ratio]{Plotting the ratio between initial and
        instantaneous temperature for varying $T_i$. Here the initial
        temperatures ranges from 20 to 500 kelvin, with 48 intermediate steps.
    As we can see the ratio stabilizes around 0.5. The graph of an arbitrary
temperature has been highlighted in black.}
    \label{fig:temperature_ratio}
\end{figure}

@@

In order to measure the melting temperature of the system, we can use the Einstein relation which states the following:
\begin{equation}
    \notag
    \langle r^2(t) \rangle = 6Dt.
\end{equation}
$D$ is here the diffusion constant, $t$ is time and the \emph{mean square displacement} is calculated as
\begin{equation}
    \notag
    r_i^2(t) = \left| \vec{r}_i(t) - \vec{r}_i(0) \right|^2
\end{equation}
for atom $i$.

\clearpage

\begin{thebibliography}{9}
\bibitem{min_image}
    \url{https://en.wikipedia.org/wiki/Periodic_boundary_conditions#Practical_implementation:_continuity_and_the_minimum_image_convention}
\bibitem{equipar}
    \url{https://en.wikipedia.org/wiki/Equipartition\_theorem}
\bibitem{virial}
    \url{https://en.wikipedia.org/wiki/Virial_theorem}
\end{thebibliography}
\end{document}
